---
title: "Analysis of Suicide and HDI Data in 2019"
author: "Adrián CALVO VALDERRAMA"
format: 
  dashboard:
    nav-buttons: 
    - icon: github
      href: https://github.com/adcalvoval/adcv_RMCQ3_capstone
      target: _blank
    - icon: clipboard2-data
      href: https://www.gapminder.org/data/
      target: _blank
    - icon: globe
      href: https://data.worldbank.org/indicator/SP.POP.TOTL?most_recent_year_desc=false
      target: _blank
theme: flatly
execute:
  echo: false
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               here,
               sf,
               bslib, 
               bsicons,
               rnaturalearth, 
               plotly, 
               countrycode, 
               htmltools, 
               reactable,
               janitor,
               readxl,
               poliscidata,
               stringr,
               formattable
               )
options(scipen = 999)
```

```{r}
# Getting the data

happy <- read_csv(here("data/hapiscore_whr.csv"))
gdp <- read_csv(here("data/gdp_pcap.csv"))
unemp_male <- read_csv((here("data/male_long_term_unemployment_rate_percent.csv")))
suicide <- read_csv(here("data/suicide_total_deaths.csv"))
geo <- read_excel(here("data/Data Geographies - v2 - by Gapminder (2).xlsx"), sheet = 2)
pop <- read_csv(here("data/pop.csv")) # missing data for significant countries (e.g. China, India)
pop2 <- read_csv(here("data/pop_wb.csv"))
hdi<- read_csv(here("data/hdi_human_development_index.csv"))
```

```{r include=FALSE}
suicide_clean <- 
  suicide %>%
  mutate(across(`1990`:`2019`, ~ str_replace_all(.x, "(\\d+\\.?\\d*)k", "\\1e3"))) %>%  
  mutate(across(`1990`:`2019`, ~ str_replace_all(.x, "(\\d+\\.?\\d*)M", "\\1e6"))) %>%   
  mutate(across(`1990`:`2019`, ~ as.numeric(.x)))  

# pop_clean <-
#   world %>% 
#   select(country, pop_total) %>% 
#   mutate(population = (pop_total * 1000000)) %>% 
#   select(!pop_total)
# 
# 
# pop_clean2 <-
#   pop %>% 
#   select(country,`1990`:`2019`) %>% 
#     mutate(across(`1990`:`2019`, ~ str_replace_all(.x, "(\\d+\\.?\\d*)k", "\\1e3"))) %>%  
#     mutate(across(`1990`:`2019`, ~ str_replace_all(.x, "(\\d+\\.?\\d*)M", "\\1e6"))) %>%   
#     mutate(across(`1990`:`2019`, ~ as.numeric(.x))) %>% 
#   pivot_longer(cols = `1990`:`2019`,
#                names_to = "year",
#                values_to = "population")


# View(pop_clean2)

pop_wb_clean <- 
  pop2 %>% 
  select(!`Indicator Name`:`Indicator Code`) %>% 
  rename(country = `Country Name`) %>% 
  mutate(countries = case_when(
    country == "United Kingdom" ~ "UK",
    country == "United Arab Emirates" ~ "UAE",
  TRUE ~ country  
)) %>% 
  select(!country) %>% 
  select(countries, everything())

View(pop_wb_clean)

setdiff(suicide_clean$country, pop_wb_clean$country)
setdiff(pop_wb_clean$country, suicide_clean$country)

suicide_clean <- 
  suicide_clean %>% 
  mutate(countries = case_when(
  country == "Bahamas" ~ "Bahamas, The",
  country == "Brunei" ~ "Brunei Darussalam",
  country == "Congo, Democratic Republic of the" ~ "Congo, Dem. Rep.",
  country == "Congo, Republic of the" ~ "Congo, Rep.",
  country == "Cape Verde" ~ "Cabo Verde",
  country == "Czech Republic" ~ "Czechia",
  country == "Egypt" ~ "Egypt, Arab Rep.",
  country == "Iran" ~ "Iran, Islamic Rep.",
  country == "Kyrgyzstan" ~ "Kyrgyz Republic",
  country == "Korea, South" ~ "Korea, Rep.",
  country == "Laos" ~ "Lao PDR",
  country == "Macedonia" ~ "North Macedonia",
  country == "Burma (Myanmar)" ~ "Myanmar",
  country == "Korea, North" ~ "Korea, Dem. People's Rep.",
  country == "Russia" ~ "Russian Federation",
  country == "Slovakia" ~ "Slovak Republic",
  country == "Swaziland" ~ "Eswatini",
  country == "Syria" ~ "Syrian Arab Republic",
  country == "Turkey" ~ "Turkiye",
  country == "Venezuela" ~ "Venezuela, RB",
  country == "Vietnam" ~ "Viet Nam",
  country == "Yemen" ~ "Yemen, Rep.",
  country == "USA" ~ "United States",
  TRUE ~ country  
)) %>% 
  select(!country) %>% 
  select(countries, everything())

View(suicide_clean)

suicide_19 <- 
  suicide_clean %>% 
  select(countries, `2019`) %>% 
  rename(suicides_19 = `2019`)


suicide_90_19 <-
  suicide_clean %>% 
  select(countries, `1990`, `2019`) %>% 
  mutate(pct_increase = round(((`2019`-`1990`)/`1990`)*100))

highest_suicide_country <- 
  suicide_19 %>% 
  arrange(desc(suicides_19)) %>% 
  head(1) %>% 
  pull(countries)

pop_wb_clean_19 <-
    pop_wb_clean %>% 
      select(countries, `2019`) %>%
      rename(pop_19 = `2019`)

View(pop_wb_clean_19)


suicide_pop_19 <-
  suicide_19 %>% 
  left_join(pop_wb_clean_19, 
            by = "countries") %>%  
  mutate(suicide_per10K = round((`suicides_19`/`pop_19`)*10000, 2))

View(suicide_pop_19)

suicide_long <-
  suicide_clean %>% 
  pivot_longer(cols = `1990`:`2019`,
               names_to = "year",
               values_to = "suicides")

highest_suicide_country_10K <- 
  suicide_pop_19 %>% 
  arrange(desc(suicide_per10K)) %>% 
  head(1) %>% 
  pull(countries)

total_suicide_greenland_19 <-
  suicide_pop_19 %>% 
  arrange(desc(suicide_per10K)) %>% 
  head(1) %>% 
  pull(suicides_19)

highest_suicide_19 <-
  suicide_pop_19 %>% 
  arrange(desc(suicides_19)) %>% 
  head(1) %>% 
  pull(suicides_19)

highest_suicide_country <-
  suicide_pop_19 %>% 
  arrange(desc(suicides_19)) %>% 
  head(1) %>% 
  pull(countries)

highest_increase_90_19_country <-
  suicide_90_19 %>% 
  arrange(desc(pct_increase)) %>% 
  head(1) %>% 
  pull(countries)

highest_increase_90_19 <-
  suicide_90_19 %>% 
  arrange(desc(pct_increase)) %>% 
  head(1) %>% 
  pull(pct_increase)


happy_clean <- 
  happy %>% 
  select(country, `2005`:`2019`) %>% 
  pivot_longer(cols = `2005`:`2019`,
               names_to = "year",
               values_to = "happindex") %>% 
  mutate(countries = 
           case_when( 
             country == "USA" ~ "United States",
             country == "Venezuela" ~ "Venezuela, RB",
             country == "Czech Republic" ~ "Czechia",
             country == "Russia" ~ "Russian Federation",
             country == "Egypt" ~ "Egypt, Arab Rep.",
             country == "Syria" ~ "Syrian Arab Republic",
             country == "Vietnam" ~ "Viet Nam",
             country == "Yemen" ~ "Yemen, Rep.",
             country == "Turkey" ~ "Turkiye",
         TRUE ~ country)) %>% 
  select(!country) %>% 
  select(countries, everything())

View(happy_clean)

suicide_long <-
  suicide_clean %>%
  pivot_longer(cols = `1990`:`2019`,
               names_to = "year",
               values_to = "suicides")

View(suicide_long)


setdiff(suicide_long$countries, happy_clean$countries)
setdiff(happy_clean$countries, suicide_long$countries)


suicide_happy <- 
  left_join(suicide_long, happy_clean, by = c("countries", "year"))

View(suicide_happy)

hdi_long <-
  hdi %>% 
  select(country, `1990`:`2019`) %>% 
  pivot_longer(cols = `1990`:`2019`,
               names_to = "year",
               values_to = "hdi")  %>% 
  mutate(countries = 
           case_when( 
             country == "USA" ~ "United States",
             country == "Russia" ~ "Russian Federation",
         TRUE ~ country)) %>% 
  select(!country) %>% 
  select(countries, everything())
  

View(hdi_long)

top_hdi <- 
  hdi_long %>%
    group_by(year) %>%
    slice_max(hdi, n = 1) %>%  
    select(year, countries, hdi)  

View(top_hdi)


setdiff(happy_clean$countries, hdi$country)
setdiff(hdi$country, happy_clean$countries)

pop_long <-
  pop_wb_clean %>% 
  select(!`Country Code`) %>% 
  select(!`1960`:`1989`) %>% 
  select(!`2020`:`2023`) %>% 
  pivot_longer(cols = `1990`:`2019`,
               names_to = "year",
               values_to = "population") %>% 
  mutate(countries = 
           case_when( 
             countries == "United Arab Emirates" ~ "UAE",
             countries == "United Kingdom" ~ "UK",
             countries == "Korea, Rep." ~ "South Korea",
             countries == "Korea, Dem. People's Rep." ~ "North Korea",
             countries == "Lao PDR" ~ "Lao",
             countries == "West Bank and Gaza" ~ "Palestine",
             countries == "Gambia, The" ~ "Gambia",
         TRUE ~ countries)) 

View(pop_long)


complete_data <-
  left_join(suicide_happy, hdi_long, by = c("countries", "year"))

View(complete_data)
```


```{r include=FALSE}

# Plots

suicide_plot <-
  suicide_long %>%
  filter(suicides > 10000) %>%
 ggplot(aes(x = year,
            y = suicides,
            frame = year,
            color = countries)) +
 geom_point() +
 theme_minimal()

suicide_ggplotly <- ggplotly(suicide_plot)



top_5_graph <-
  suicide_long %>%
  filter(countries %in% c("Lesotho", "Guyana", "Suriname", "Greenland", "Lithuania")) %>%  
      ggplot() +
  geom_line(aes(x = year,
                y = suicides,
                group = countries,
                colour = countries)) + 
  theme_grey()

top_5_graph

top_5_countries <- 
  suicide_19 %>% 
  arrange(desc(suicides_19)) %>% 
  head(5) %>% 
  mutate(tooltip_label = paste(countries, 
                               suicides_19, 
                               sep = ": ")) %>% 
  ggplot(aes(y = reorder(countries, suicides_19), 
             x = suicides_19, 
             fill = suicides_19, 
             text = tooltip_label)) + 
  geom_col() +  
  labs(y = "Country", 
       x = "Number of Suicides") + 
  theme(legend.position = "none") 


top_5_countries_ggplotly <- 
  ggplotly(top_5_countries, tooltip = "text")

top_5_countries_ggplotly



top_5_countries_10k <- 
  suicide_pop_19 %>% 
  arrange(desc(suicide_per10K)) %>% 
  head(5) %>% 
  mutate(tooltip_label = paste(countries, 
                               suicide_per10K, 
                               sep = ": ")) %>% 
  ggplot(aes(y = reorder(countries, suicide_per10K), 
             x = suicide_per10K, 
             fill = round(suicide_per10K, 2), 
             text = tooltip_label)) + 
  geom_col() +  
  labs(y = "Country", 
       x = "Number of Suicides per 10,000 people") + 
  theme(legend.position = "none") 




top_5_countries_10k_ggplotly <- 
  ggplotly(top_5_countries_10k, tooltip = "text")

top_5_countries_10k_ggplotly
```


```{r include=FALSE}

# Mapping

country_shapes <- rnaturalearth::ne_countries()

setdiff(suicide_pop_19$countries, country_shapes$sovereignt)
setdiff(country_shapes$sovereignt, suicide_pop_19$countries)


suicide_geo <- 
  suicide_pop_19 %>% 
     mutate(country = case_when(
        countries == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
        countries == "Congo, Rep." ~ "Republic of the Congo",
        countries == "Tanzania" ~ "United Republic of Tanzania",
        countries == "Bahamas, The" ~ "The Bahamas",
        countries == "Russian Federation" ~ "Russia",
        countries == "Timor-Leste" ~ "East Timor",
        countries == "Venezuela, RB" ~ "Venezuela",
        countries == "Cote d'Ivoire" ~ "Ivory Coast",
        countries == "Eswatini" ~ "eSwatini",
        countries == "UAE" ~ "United Arab Emirates",
        countries == "Lao" ~ "Laos",
        countries == "Viet Nam" ~ "Vietnam",
        countries == "Kyrgyz Republic" ~ "Kyrgyzstan",
        countries == "Iran, Islamic Rep." ~ "Iran",
        countries == "Syrian Arab Republic" ~ "Syria",
        countries == "Turkiye" ~ "Turkey",
        countries == "Brunei Darussalam" ~ "Brunei",
        countries == "Slovak Republic" ~ "Slovakia",
        countries == "Yemen, Rep." ~ "Yemen",
        countries == "Egypt, Arab Rep." ~ "Egypt",
        countries == "Serbia" ~ "Republic of Serbia",
        TRUE ~ countries
      )) %>% 
          select(!country) %>% 
          select(countries, everything())

View(suicide_geo)


setdiff(suicide_geo$countries, country_shapes$sovereignt)

suicide_geo <- 
  suicide_geo %>% 
  mutate(
    country_code = countrycode(countries, 
                               "country.name", 
                               "iso3c"))

View(suicide_geo)

suicide_map <- 
  left_join(country_shapes, suicide_geo,
          by = c("adm0_a3" = "country_code")) %>% 
  mutate(tooltip_label = paste(countries,
                               round(suicide_per10K, 2), 
                               sep = ": ")) %>% 
  ggplot() + 
  geom_sf(aes(fill = suicide_per10K, text = tooltip_label)) + 
  theme_void()+ 
  theme(legend.position = "none")

suicide_map_ggplotly <- ggplotly(suicide_map, tooltip = "text") 
```

```{r include=FALSE}

# Dynamic plots

happy_suicide_plot <- 
  suicide_happy %>% 
  filter(year > 2005) %>% 
  filter(countries %in% c("Lesotho", "Ukraine", "Lithuania", "Guyana", "Greenland")) %>% 
 ggplot(aes(x = suicides, 
            y = happindex, 
            frame = year,
            color = countries)) +
 geom_point() +
 scale_x_log10() +
 theme_minimal() +
  labs(x = "Suicides",
       y = "Happiness Index",
       colour = "Countries")
  
happy_suicide_ggplotly <-
  ggplotly(happy_suicide_plot)

happy_suicide_ggplotly

complete_plot_10k <- 
  complete_data %>% 
  filter(year > 2005) %>% 
  filter(countries %in% c("Lesotho", "Ukraine", "Greenland", "Guyana", "Lithuania")) %>% 
 ggplot(aes(x = suicides, 
            y = hdi, 
            frame = year,
            color = countries)) +
 geom_point() +
 theme_minimal() +
  labs(x = "Suicides",
       y = "Human Development Index",
       colour = "Countries")


complete_plot_top <- 
  complete_data %>% 
  filter(year > 2005) %>% 
  filter(countries %in% c("United States", "Russian Federation", "India", "China", "Japan")) %>% 
 ggplot(aes(x = suicides, 
            y = hdi, 
            frame = year,
            color = countries)) +
 geom_point() +
 scale_x_log10() +
 theme_minimal() +
  labs(x = "Suicides",
       y = "Human Development Index",
       colour = "Countries")

plot_top_hdi <- 
  complete_data %>% 
  filter(year > 2005) %>% 
  filter(countries %in% c("United States", "Switzerland", "Norway")) %>% 
 ggplot(aes(x = suicides, 
            y = hdi, 
            frame = year,
            color = countries)) +
 geom_point() +
 scale_x_log10() +
 theme_minimal() +
  labs(x = "Suicides",
       y = "Human Development Index",
       colour = "Countries")


hdi_suicide_uae <- 
  complete_data %>% 
  filter(year > 2005) %>% 
  filter(countries == "UAE") %>% 
 ggplot(aes(x = suicides, 
            y = hdi, 
            frame = year,
            color = countries)) +
 geom_point() +
 scale_x_log10() +
 theme_minimal() +
  labs(x = "Suicides",
       y = "Human Development Index",
       colour = "Country")

complete_10k_ggplotly <- ggplotly(complete_plot_10k)

complete_top_ggplotly <- ggplotly(complete_plot_top)

top_hdi_ggplotly <- ggplotly(plot_top_hdi)

uae_ggplotly <- ggplotly(hdi_suicide_uae)
```



# HOME

## Row 1 {height="25%"}

```{r}
value_box(
  title = "Country with highest number of suicides in 2019",
  value = tags$p(highest_suicide_country, style = "font-size: 200%;"),
  showcase = bsicons::bs_icon("geo"),
  theme = value_box_theme(bg = "#214773"),
  p(paste0("Number of suicides in 2019: ", highest_suicide_19))
)
```

```{r}
value_box(
  title = "Country with highest number of suicides per 10,000 people in 2019",
  value = tags$p(highest_suicide_country_10K, style = "font-size: 200%;"),
  showcase = bsicons::bs_icon("geo-fill"),
  theme = value_box_theme(bg = "#518fd6"),
  p(paste0("Number of suicides in 2019: ", total_suicide_greenland_19))
)
```

```{r}
value_box(
  title = "Country with highest relative increase in the reported number of suicides",
  value = tags$p(highest_increase_90_19_country, style = "font-size: 200%;"),
  showcase = bsicons::bs_icon("graph-up-arrow"),
  theme = value_box_theme(bg = "#701515"),
  p(paste0("Increase between 1990 and 2019: ", highest_increase_90_19, " %"))
)
```

## Row 2 {height="75%"}

###  {width="60%"}

```{r title = "Number of suicides per 10,000 people per country in 2019"}
suicide_map_ggplotly 
```

###  {width="40%"}

####  {height="50%"}

```{r title = "Top 5 countries: number of suicides in 2019"}
top_5_countries_ggplotly
```

####  {height="50%"}

```{r title = "Top 5 countries: number of suicides per 10,000 people in 2019"}
top_5_countries_10k_ggplotly
```

# PLOTS

## {height="50%" .flow}

### {width="50%"}

```{r title = "Evolution of suicide rates and HDI in countries with highest number of suicides per 10,000 people"}
complete_10k_ggplotly
```

### {width="50%"}

```{r title = "Evolution of suicide rates and HDI in countries with highest number of suicides"}
complete_top_ggplotly
```


## {height="50%" .flow}

### {width="50%"}

```{r title = "Evolution of suicide rate and Happiness Index in countries with highest number of suicides per 10,000 people"}
happy_suicide_ggplotly
```

### {width="50%"}

```{r title = "Evolution of suicide rate and HDI in the countries with highest HDI" }
top_hdi_ggplotly
```



# DATA

The data used in this dashboard is shown below and can be downloaded as a CSV.

```{r}
library(htmltools)

htmltools::browsable(
  tagList(
    reactable(complete_data, 
              elementId = "complete_data", 
              searchable = T, 
              filterable = T), 
    
tags$button("Download as CSV", 
            onclick = "Reactable.downloadDataCSV('complete_data')")
  )
)
```

# ABOUT

This data comes from the r `gapminder` package, and is originally sourced from the Gapminder Foundation.

The Gapminder Foundation is a non-profit venture registered in Stockholm, Sweden, that promotes sustainable global development and achievement of the United Nations Millennium Development Goals by increased use and understanding of statistics and other information about social, economic, and environmental development at local, national, and global levels.

Gapminder was founded in 2005 by Ola Rosling, Anna Rosling Rönnlund, and Hans Rosling. The name Gapminder was derived from the "Mind the Gap" warning messages on the London Underground.

An example of one of Hans Rosling's videos is shown below:

<iframe width="560" height="315" src="https://www.youtube.com/embed/hVimVzgtD6w?si=hjxf5ryxh7mx5A4R" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen>

</iframe>
